Imputation of Incomplete Multilevel Data with R

### 2023-12-12 docs/manuscript.tex
Integrate PR #42

\begin{tcolorbox}[enhanced jigsaw, breakable, colback=white, opacityback=0, left=2mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, rightrule=.15mm, leftrule=.75mm]

@@

When imputing clustered data, we initially focus on the research
question and the intended analysis, assuming the presence of
non-incomplete values. This assessment sheds light on research
hypotheses and the contemplated main(s) statistical model(s) and it
provides insights into the data's structure (refer to the level table),
variable types (e.g., confounders, auxiliary variables), and other
considerations about variables relationships such as interactions or
polynomial terms.

@@

Next, it is necessary to explore the information available in the data.
Exploring individual variables using tools such as histograms or QQ
plots can help to delineate variable distributions and plausible ranges
of values and also identify input errors or outliers. Evaluating
interactions between predictors using scatter plots or conditional
plots, assessing collinearity using VIF or correlations, can help to
glimpse nonlinear relationships that may affect the original formulation
of the main model. Further testing of assumptions related to the
response variable, including variance (homogeneity using conditional box
plots), independence (e.g., ACF, variograms) and response vs.~predictor
relationships (conditional plots - x vs.~y), may serve to choose an
imputation model more suitable to the data. In addition, the intraclass
correlation coefficient (ICC) can be examined to assess cluster
differences, aiding in the choice between the 2l and 1l methods for
imputation.

One can also explore missingness, specifically missing patterns at the
cluster level to identify types of patterns (e.g., non-monotonic,
systematic) to guide the selection of an appropriate imputation approach
in terms of computational efficiency (e.g., simpler regression
imputation versus FCS in univariate patterns). Similarly, these missing
patterns can also be used to identify potential predictors in individual
imputation models by means of inflow criteria (quantifying the
connections between missing data in one variable and other observed
variables) and outflow criteria (determining the connections between
observed values in one variable and missing data in other variables).
Although several packages exist for visualizing missing data, this
tutorial focuses on the recently implemented ggmice package.

@@

Prior to implementing multiple imputation, it is critical to assess the
robustness of the estimation model when confronted with missing data.
There are scenarios in which specific Maximum Likelihood (ML) estimation
methods outperform Multiple Imputation (MI) methods. For instance, when
the response variable is the sole incomplete variable, mixed models
demonstrate robustness to missing data under the Missing at Random (MAR)
assumption and with a correct variance-covariance
specification.\cite{Molenberghs_2007}. Furthermore, depending on the
proportion of missingness (usually is \textless5\%), opting for a
simpler complete case analysis might be sufficient.

@@

Before proceeding with the imputation model, it is necessary to filter
the dataset to include only the minimal variables required for the main
model. If additional procedures are conducted during the analysis, it is
essential to include variables associated with these procedures, such as
confounders in the case of balancing techniques. Other relevant
variables, like instrumental variables or auxiliary variables, may
enhance parameter estimates even if not included in the main model, as
they could be linked to the probability of missingness for some
incomplete variables.

Furthermore, at the variable level, it is crucial to assess whether
proxy variables should be considered or if deterministic imputation
alone is adequate for imputing missing values. For instance, certain
incomplete variables may not necessitate stochastic imputation methods
like MICE. Instead, they can be effectively addressed through deductive
imputation, where incomplete values are inferred from logical and
deterministic relationships between variables. This approach is
especially beneficial for variables that are functions of others, such
as deriving a person's BMI from their weight and height. It also proves
valuable for determining values for one-level variables from two-level
ones, as seen in the context of Individual Participant Data (IPD). In
such cases, incomplete information can be deduced from metadata, like
inferring incomplete data about abortion in a country where abortion is
illegal, or through cross-temporal or protocol-based deduction, such as
imputing missing test values for deceased patients.

@@

Various strategies can be employed in the imputation process to account
for clustering, including the inclusion of a cluster indicator variable,
conducting a separate imputation process for each cluster, or employing
a simultaneous imputation method that considers clustering\cite{eddings}

The choice of strategy depends primarily on the assumptions made in the
main analysis and the constraints imposed by the analyzed data.
Concerning data restrictions when the analysis do not contemplate an
hierarchical model (eg. descriptive way), if there are few clusters with
many observations per cluster, the cluster indicator or imputation
separated by groups may be appropriate \cite{graham2009}. Conversely, an
imputation model that simultaneously imputes all clusters using a
hierarchical model can be employed when there are more clusters or fewer
observations per cluster \cite{Allison_2002}.

In a hierarchical imputation model, correlations between observations
within clusters are modeled by a random effect, allowing estimation even
with a limited number of observations per cluster. Multiple imputation
models based on hierarchical models have been proposed, each with
different assumptions. \cite{audigier}

In the particular case that the analysis contemplates the use of a
hierarchical model (i.e.~main model), before choosing a particular
strategy, it is crucial to verify if the assumptions of the imputation
model align with those of the main analysis. For example, using the
cluster indicator strategy may introduce bias estimates when the model
is based on a hierarchical structure \cite{taljaard2008,speidel2018}.
Even if an imputation strategy congruent with the main model is
preferred, it is essential to assess its appropriateness for the data,
considering that less complex imputation strategies may lead to unbiased
estimates in certain scenarios \cite{bailey2020}.

@@

The initial step involves selecting the individual imputation model for
each incomplete variable in the dataset. The mice package automatically
proposes imputation methods based on the variable type for 1l variables.
For 2l level variables, the micemd package's () function can be
utilized. This function selects among the 2l imputation methods based on
the size of the clusters and the proportion of missingness in each
cluster.

In addition to the package-defined imputation methods, users can specify
custom methods using the ``I formula''. This flexibility allows for the
calculation of derived variables internally or adjustments to imputation
methods based on specific conditions, such as conditioning the
imputation model on the level of an incomplete covariate (e.g.,
pregnancy test on females).

@@

The imputation model can include additional terms than the main model
terms that do not lead to uncongeniality. For instance, it is advisable
to include the outcome variable in the imputation model for prediction
variables \cite{moons2006a}. In cases where the outcome is
time-to-event, the Nelson-Aalen estimate of the time to the event should
be included as a covariate in the imputation model {[}REF{]}.
Furthermore, the inclusion of auxiliary variables, even if not part of
the main model, can be associated with the probability of missingness,
enhancing the likelihood of satisfying the Missing at Random (MAR)
assumption and improving estimation efficiency \cite{hardt2012a}.

The specification of imputation models is done on a variable basis,
either using a prediction matrix where the type of prediction variable
should be specified accordingly (see table) or through a list of
formulas in the formula parameter. The latter option proves useful in
formulating imputation models with polynomial terms or interactions
compared to the prediction matrix specification, which requires the
inclusion of additional terms as Just Another Variable (JAV)
\cite{buuren2018a}.

However, even when considering an interaction term, depending on the
adapted individual imputation model, it has been suggested, for
instance, for treatment interaction effects, to conduct separated
imputation by treatment group \cite{zhang2023}. Additionally, there are
imputation models based on random forest or deep learning that can
handle interaction and non-linear terms that do not require the explicit
inclusion of the non-linear terms.

@@

@@

>>>>

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, arc=.35mm, rightrule=.15mm, breakable, opacityback=0, bottomrule=.15mm, colback=white, leftrule=.75mm, left=2mm]

@@

When dealing with incomplete clustered data, start by looking at your
research questions and planned analysis. Pretend there are no missing
data at first. This will help you understand your research hypotheses,
the main statistical model, and give you insights into your data's
structure (refer to the level table), variable types (e.g., confounders,
auxiliary variables), and other considerations such as variable
relationships, interactions or polynomial terms.

@@

Now, explore what's in your data. Use simple tools like histograms and
QQ plots to understand variable distributions, plausible range of values
and identify errors or outliers. Check interactions between predictors
with scatter plots, look for collinearity with VIF or correlations
plots, this may also help you to identify non-considered relationships
that may affect the main model. You might also consider to test
assumptions related to your response variable such as variance
homogeneity, independence between observations (eg. ACF, variagrams).
This will help you choose an imputation model that suits your data. In
addition, the intraclass correlation coefficient (ICC) can be examined
to assess cluster differences, aiding in the choice between the 2l and
1l methods for imputation.

Next, explore the missingness. Look at the
\textbf{proportion of missing values} in the dataset variables. This
helps find potential predictors and cut down on unnecessary variables in
the imputation model. Doing this can lower the risk of multicollinearity
or computational issues, especially with certain parametric imputation
methods. You can also identify predictors for the imputation model by
using inflow criteria to see connections between missing data in one
variable and observed variables, and outflow criteria to identify
connections between observed values in one variable and missing data in
others.

Check the missing patterns at cluster level, this can help you to select
the most appropiated imputation approach in terms of computational
efficiency (e.g., simpler regression imputation versus FCS in univariate
patterns).

@@

Before diving into imputation, make sure your estimation model can
handle missing data. Sometimes, simpler methods like complete case
analysis might be suffice, especially if your missingness is low
(usually \textless5\%). There are scenarios in which specific Maximum
Likelihood (ML) estimation methods outperform Multiple Imputation (MI)
methods, for instance when the response variable is the sole incomplete
variable, mixed models demonstrate robustness to missing data under the
Missing at Random (MAR) assumption and with a correct
variance-covariance specification.\cite{Molenberghs_2007}.

@@

Clean up your dataset before the imputation process. Keep initially the
essential variables for your main model, but include extra variables if
needed for specific procedures during analysis (e.g.~confounders on
balancing procedures). Think about adding other useful variables, even
if they're not in the main model, such as instrumental or auxiliary
variables which might boost your parameter estimates, especially if
they're associated to the probability of missingness for some incomplete
variables.

Figure out if you can directly impute incomplete variables by just use
deductive imputation. This involves inferring missing values based on
logicallogical connections between variables. It's especially handy for
variables that depend on each other, like calculating BMI from weight
and height.

Deductive imputation is also useful for getting values for level-1
variables from level-2 ones, like in Individual Participant Data (IPD).
In these situations, you can guess missing information from metadata or
by using or through deduction based on time or protocols. For example,
you could deduce missing test values for patients who have passed away.

@@

When it comes to handling clusters during the imputation process, you've
got a few options. You can use a cluster indicator variable, run
separate imputation for each cluster, or go for a simultaneous
imputation method that takes clustering into account \cite{eddings}.

Which strategy you choose depends on the assumptions in your main
analysis and the limitations of your data. If your analysis do not use a
hierarchical model (like a descriptive approach) and you have a small
number of clusters with lots of observations in each, using a cluster
indicator or separate imputation might be the way to go
\cite{graham2009}. Conversely, if you have more clusters or fewer
observations per cluster, you might want to try a simultaneous
hierarchical imputation model \cite{Allison_2002}.

In a hierarchical imputation model, random effects model the
correlations between observations within clusters, making it possible to
estimate even with a small number of observations per cluster. There are
various proposed multiple imputation models based on hierarchical
models, each with its own set of assumptions \cite{audigier}.

If your analysis uses a hierarchical model, make sure the assumptions of
your imputation model match up. For example, using a cluster indicator
approach may lead to bias estimates if your model is based on a
hierarchical structure \cite{taljaard2008,speidel2018}. Even if you
prefer an imputation strategy that aligns with your main model, check if
it suits your data; sometimes simpler strategies can give unbiased
estimates in certain scenarios \cite{bailey2020}.

@@

Start by choosing the imputation model for each incomplete variable in
your dataset. The mice package suggests methods based on variable types
for non-clustered variables, and for the cluster ones, you can use the
micemd package's find.defaultMethod() function. This function selects
from different 2l imputation methods based on cluster size and the
proportion of missing data in each cluster.

Besides the package-defined imputation methods, you can specify custom
methods using the ``I formula''. This lets you calculate deterministic
variables during the imputation or tweak imputation methods based on
specific conditions, like conditioning the imputation model to the level
of an incomplete covariate (e.g., a pregnancy test for females).

@@

The imputation model can incorporate additional terms compared to the
main model without causing compatibility issues. For instance, it's
recommended to include the outcome variable in the imputation model for
prediction variables \cite{moons2006a}. In cases where the outcome is
time-to-event, the Nelson-Aalen estimate of the time to the event should
be added as a covariate in the imputation model {[}REF{]}. Additionally,
including auxiliary variables, even if not part of the main model, can
be linked to the probability of missingness, improving the likelihood of
meeting the Missing at Random (MAR) assumption and enhancing estimation
efficiency \cite{hardt2012a}.

Imputation models are specified on a variable basis, either using a
prediction matrix in the pred parameter or through a list of formulas in
the formula parameter. In the prediction matrix option, the type of each
predictor variable is specified for each incomplete variable (see
table).

\begin{align}
y_{ij} =& (\beta_0 + b_{0_i})+ \beta_1x_{1_{ij}}+\beta_2x_{2_{i.}} \nonumber \\
+& (\beta_3 +  b_{3_i})x_{3_{ij}} \nonumber \\
+& \beta_4x_{4_{ij}} + \beta_{m4}\overline{x_{4_{i.}}}\nonumber \\
+& (\beta_5 +  b_{5_i})x_{5_{ij}} +\beta_{m5}\overline{x_{5_{i.}}}\nonumber \\
\end{align}

\begin{tabular}{ll}
     \toprule
Type & Definition \\
  \midrule
-2  & Cluster variable, in this case the one defined by j index  \\
 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
 2  & Random variable, e.g., $x_{3_{ij}}$\\
 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
-3. & Random variable only included on selection model (Heckman model)\\
-4. & Random variable only included on main model (Heckman model)\\
      \bottomrule
\end{tabular}

\begin{itemize}
\tightlist
\item
  3 and 4 type have been advised to used as the inclusion of the means
  of the cluster is beneficial on FCS \cite{mistler2017}
\end{itemize}

Recipes have been proposed for imputing incomplete level-1 and level-2
variables for hierarchical models \cite{buuren2018a} (\(\S\) 7.10),
which can be convenient to follow when dealing with numerous variables
(interaction terms at different levels) and models with many random
effects that are prone to convergency problems due to overspecification
in the imputation model. These rules were designed to ensure
compatibility among the conditionally specified imputation models and
congeniality between the imputation and the main model. However, they
may not be applicable to all 2l imputation methods; for instance, the
2l.2stage imputation methods of the micemd package only allow the
inclusion of random predictor variables (2).

On the other hand, the formula option is useful in specifying complex
imputation models with polynomial terms or interactions and compared
with the prediction matrix method do not requires the inclusion of
additional terms as Just Another Variable (JAV)
\cite{buuren2018a}(\(\S\) 6.4).

For some interaction terms, it has been suggested, for instance, for
treatment interaction effects, to conduct separate imputation by
treatment group \cite{zhang2023}. Additionally, it can be used
imputation models based on random forest or deep learning that can
handle interaction and non-linear terms without requiring the explicit
specification of an imputation model.

@@

Given the \(j\)-th student belonging to the \(i\)-th school, the main
model can be formulated as:

\[ popular_{ij} = (\beta_0 + b_i) + \beta_1sex_{ij}+ \beta_2texp_{ij}+\epsilon_{ij}\]
where \(\epsilon_{ij}\) corresponds to the error term.

@@

In this dataset all the variables are level-1 (patient), except by the
level-2 (study) type variable. The analysis model for this dataset is a
prediction model with \texttt{mort} as the outcome. In this tutorial
we'll estimate the adjusted prognostic effect of \texttt{ct} on
mortality outcomes. The estimand is the adjusted odds ratio for
\texttt{ct}, after including \texttt{type}, \texttt{age}
\texttt{motor\_score} and \texttt{pupil}. Therefore the main model for
the \(i\)-th patient from the \(j\)-th study can be described as:

\[ mort_{ij} = (\beta_0 + b_i) + \beta_1type_{.j}+ \beta_2age_{ij}+ \beta_3motorscore_{ij} + \beta_4pupil_{ij} + \beta_5ct_{ij}+\epsilon_{ij}\]
where \(\epsilon_{ij}\) corresponds to the error term.
